---
title: "Exploration"
format: gfm
---

This document is for me to explore different aspects of the data before creating the dashboard. `tidyverse` libraries will be used to assist in exploring the data.

```{r}
#| message: false
library(tidyverse)
```


# `README`
Now BART's data contains a `README` file. This mostly gives an overview of the column names and how they are stored. The columns are:

```{r}
col_names <- c(
  "day",
  "hour",
  "origin_station",
  "destination_station",
  "trip_count"
)
```

These are stored to help with reading the data later.

The file also mentions that the names of the different stations are abbreviated and provides the [download link for a spreadsheet](https://www.bart.gov/sites/default/files/docs/Station_Names.xls) giving the corresponding full names. After downloading, I was able to read this into a tibble.

```{r}
station_tibble <- readxl::read_xls("Station_Names.xls") %>%
  rename(
    station_code = `Two-Letter Station Code`,
    station_name = `Station Name`
  )

station_tibble
```

The full `README` can be found [here](http://64.111.127.166/origin-destination/READ%20ME.txt).

# Reading Data
I decided to preview the most recent data.

```{r}
#| message: false
read_csv(
  "http://64.111.127.166/origin-destination/date-hour-soo-dest-2022.csv.gz",
  col_names = col_names
) %>%
  head()
```

Now, one issue with this data is that the station abbreviations do not match those given in the spreadsheet.

